Variance reduction for Markov chains via martingale representations

### 2020-05-04 aap/Revision/vr-mcmc-revision-final.tex
update

\paragraph{Notations. }\label{par:notations}
We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution. For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm. For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$. For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
\[
\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,y_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
\]
In the particular case $m=1$ we drop the subscript $y_1$ in that notation. For probability measures $\mu$ and $\nu$ on $\rset^d$ denote by $\tvnorm{\mu-\nu}$ the total variation distance between $\mu$ and $\nu$, that is,
\[
\tvnorm{\mu-\nu}=\sup_{A \in \mathcal{B}(\rset^d)}
|\mu(A)-\nu(A)| \eqsp.
\]
where $\mathcal{B}(\rset^d)$ is a Borel $\sigma$-algebra of $\rset^d$. For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$. Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} |f(x)|/V(x)$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
\[
d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\nolimits_{\rset^d}f(x)\,d\mu(x) - \int\nolimits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
\]

@@

\hat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2

@@

\hat{a}_{r,k}(x)= \int \hat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z) \quad \text{and}

@@

\hat{A}_{q,k}(x) = \sum_{r=0}^q \hat{a}_{r,k}(x) \eqsp,

@@

and construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form

@@

\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f)

@@

In the next section we analyze the case of ULA algorithm and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)

@@

\PE[A^2_{l,k}(X^x_{N+l-1})]\leq \rme^{\varkappa n\gamma^2}R^2_k ,\quad k \in \nset, \quad l=0,\ldots,n-1,

@@

for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  not depending  on $\gamma$ and $n$ (see \eqref{eq:Aqk2}).  Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to

@@

\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C \rme^{\varkappa n\gamma^2}}{n}.

@@

 Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \rme^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi.

@@

Our analysis is carried out under the following two assumptions:

@@

\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.

@@

\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds

@@

\ps{D^2U(x)}{x} \geq m\|x\|^2.

@@

@@

\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y

@@

$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias. Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017},

@@

 V(x) = 1+ \|x\|^2 \,.

@@

@@

Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ and $x \in \rset^d$,

@@

\PVar\left[\pi_n^N(f)\right]\leq \frac{C}{n\gamma} + C\left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2 \eqsp.

@@

X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)

@@

Fix some $K\geq 1$ and suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable

@@

Then, there exist $\gamma_0 > 0$, $C_K < \infty$, and $\varkappa > 0$  such that, for all $\gamma \in (0,\gamma_0]$, $n \in \nset$, and $x \in \rset^d$,

@@

A^2_{q,\mathbf{k}}(x)\leq C_K \rme^{\varkappa n\gamma^2}\{1+V(x)\} \,,\quad 1\leq |\mathbf{k}|\leq K, \quad \text{for all  $q=0,\ldots,n-1,$}

@@

\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \rme^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma}{2}\right)^{|I|K-1} \, \{1+V(x)\} \eqsp,

@@

Under assumptions of Theorem~\ref{th:mr}, there exist $\gamma_0 > 0$, $\tilde{C} < \infty$ and $\varkappa > 0$ such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$, $n\in \nset$, and $x \in \rset^d$,

@@

\PVar\left[\pi_{K,n}^{N}(f)\right] \leq \tilde{C}  n^{-1} \gamma^{K-1}  \rme^{\varkappa n\gamma^2}
\eqsp.

@@

@@

Finally, using the Gaussian Poincare inequality, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds

@@

\right)\leq C_{K}\gamma^{-1}\rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=0,\ldots,n-1,

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic \eqref{eq:v_ergodicity} with $V(x) = 1+\|x\|^2$ and $\rho$ specified in Lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:var_bound},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
Now from Taylow expansion of denominator it follows that $1-\rho^{1/2} \sim C\gamma$ with constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2
\]

@@

\right)\leq C\gamma^{-1} \rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,

@@

\begin{lem}
\label{lem:var_poincare}
Under assumptions of Theorem~\ref{th:mr}, it holds
$$
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C \rme^{\varkappa n\gamma^2}(1+V(x)), \quad q=1,\ldots,n,
$$
where $C,\varkappa>0$ are constants that do not depend on $n$ and $\gamma.$
\end{lem}
\begin{proof}
The expression
$\sum_{p=1}^q \partial_{y_1}^K f(X_p^x)$
can be viewed as a deterministic function of
$x,Z_1,Z_{2},\ldots,Z_q$
$$
\sum_{p=1}^q  \partial_{y_1}^K f(X_p^x)
=F(x,Z_1,Z_{2},\ldots,Z_q).
$$
By the (conditional) Gaussian Poincar\'e inequality,
we have
$$
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
\le\mathsf E_x\left[
\|\nabla_Z F(x,Z_1,Z_{2},\ldots,Z_q)\|^2
\right],
$$
where $\nabla_Z F=(\partial_{Z_1} F,\ldots,\partial_{Z_q} F)$,
and $\|\cdot\|$ denotes the Euclidean norm.
Notice that \(\partial_{Z_j} F=\sqrt{\gamma}\,\partial_{y_j} F\) and
hence
$$
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=1}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
$$
It is straightforward to check that
$\partial_{y_j}\partial_{y_1}^K f(X_p^x)=0$
whenever $p<j$. Therefore, we get
\begin{equation}\label{eq:10062018a2}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=j}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
\end{equation}
Now fix $p$ and $j$, $p\ge j$, in $\{1,\ldots,q\}$.
By Fa\`a di Bruno's formula
\[
\partial_{y_{1}}^{K}f\left(X_{p}^x\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p}^x)\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}},
\]
where the sum is taken over all $K$-tuples of nonnegative integers
$(m_1,\ldots,m_K)$ satisfying
$$
1\cdot m_{1}+2\cdot m_{2}+\ldots+K\cdot m_{K}=K.
$$
Then
\begin{eqnarray*}
\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)
&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X^x_{p})\left[\partial_{y_{j}}X^x_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}^x}{k!}\right)^{m_{k}} \\
&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X^x_{p})
\sum_{s=1}^{K}\frac{m_{s}}{s!}
\left(\frac{\partial_{y_{1}}^{s}X^x_{p}}{s!}\right)^{m_{s}-1}
\\
&& \times\left[\partial_{y_{j}}\partial_{y_{1}}^{s}X^x_{p}\right]
\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}}.
\end{eqnarray*}
Using the bounds of
Lemmas \ref{lem:06062018a2} and~\ref{lem:06062018a3},
we obtain
\begin{eqnarray}
\label{eq:part-part}
\left|\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right|
\leq A_{K}\prod_{k=2}^{p}\alpha_{k}
\end{eqnarray}
with a suitable constant $A_{K}$.
Substituting this in~\eqref{eq:10062018a2},
we proceed as follows
\begin{align*}
\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
&\le
\gamma^2 A_{K}^{2}\sum_{j=1}^{q}
\mathsf{E}\left(\sum_{p=j}^{q}\prod_{k=2}^{p}\alpha_{k}\right)^{2}
\\
&\le
\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^2}
\mathsf{E}\sum_{j=1}^{q}
\left(\sum_{p=j+1}^{q+1}  \prod_{k=2}^{p} \alpha_{k}
\right)^{2}
\\
%&=
%\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
%\sum_{j=l}^{q} \gamma_{j}
%\prod_{k=l+1}^{j} \alpha_{k}^2
%\left(
%\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
%\right)^{2}
%\\
&\le
\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^3}
\mathsf{E}\sum_{j=1}^{q}
\prod_{k=1}^{j} \alpha_{k}
\left(\sum_{p=j+1}^{q+1}  \prod_{k=j+1}^{p} \alpha_{k}
\right)^{2}
\end{align*}

>>>>

\paragraph{Notations. }\label{par:notations}
We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution. For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | |y - x| < r\}$ where $|\cdot|$ is a standard Euclidean norm. For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$. For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
\[
\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,y_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
\]
In the particular case $m=1$ we drop the subscript $y_1$ in that notation. For probability measures $\mu$ and $\nu$ on $\rset^d$ denote by $\tvnorm{\mu-\nu}$ the total variation distance between $\mu$ and $\nu$, that is,
\[
\tvnorm{\mu-\nu}=\sup_{A \in \mathcal{B}(\rset^d)}
|\mu(A)-\nu(A)| \eqsp.
\]
where $\mathcal{B}(\rset^d)$ is a Borel $\sigma$-algebra of $\rset^d$. For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$. Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} \{|f(x)|/V(x)\}$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
\[
d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\nolimits_{\rset^d}f(x)\,d\mu(x) - \int\nolimits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
\]

@@

\widehat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2, \quad r=0,\ldots,n-1,

@@

\widehat{a}_{r,k}(x)= \int \widehat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z),

@@

\widehat{A}_{q,k}(x) = \min\left\{W_k,\sum_{r=0}^q \widehat{a}_{r,k}(x)\right\},

@@

where $(W_k)$ is a sequence of truncation levels. Finally, we  construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form

@@

\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f).

@@

In the next section we analyze the case of ULA algorithm and show that under some  mild conditions (Lipschitz continuity and convexity outside a ball of the potential $U$),

@@

\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k ,\quad k \in \nset, \quad l=0,\ldots,n-1,

@@

for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  not depending  on $\gamma$ and $n$ (see \eqref{eq:Aqk2}).  Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:definition-hat-a-A}, then $\widehat A_{n-l,k}$ converges to

@@

\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}.

@@

% Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C  K \gamma \, T^2 \mathtt{D}_\Psi.

@@

Our analysis is carried out under the following two assumptions.

@@

| \nabla U(x)-\nabla U(y) | \leq L |x-y|, \quad x,y\in \rset^d.

@@

\item[(H2)][\textbf{Convexity}] The potential $U$ is of the form $U(x)=U_0(x)+\Delta(x),$ where $\Delta$ has  compact support and $U_0\in C^2(\mathbb{R}^d)$ with

@@

\mathfrak{m} |x|^2 \leq \ps{D^2U_0(x)}{x} \leq \mathfrak{M} |x|^2,\quad   x\in \mathbb{R}

@@

for some constants $\mathfrak{m},\mathfrak{M}>0.$

@@

\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} | y - x + \gamma \nabla U(x)|^2\right\} \rmd y

@@

$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias. Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$ which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017},

@@

 V(x) = 1+ |x|^2 \,.

@@

 The following lemma follows from Lemma~\ref{eq:cov_v_ergodic} in Appendix.

@@

Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$,

@@

\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.

@@

X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).

@@

Fix some $K\geq 1$ and suppose  that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable

@@

Moreover assume that 
\label{eq:main-ass-prod}
\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi,\quad r=2,4,
for some $\Xi=\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 
Then, there exists a constant $C_K < \infty$  such that

@@

A^2_{q,\mathbf{k}}(x)\leq C_K \,\Xi(x) \,,\quad 1\leq |\mathbf{k}|\leq K, \quad q=0,\ldots,n-1,

@@

\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma}{2}\right)^{|I|K-1} \, \Xi(x) \eqsp,

@@

Suppose that  {\bf (H2)} holds with  $\Delta \in C^2(\mathbb{R}^d)$ satisfying $|\mathsf{D}^{2}\Delta(x)|\leq D$ for all $x\in \mathbb{R}^d$ and some constant $D>0.$ Then for  $\gamma  <1/\mathfrak{M}$ we have
\sum_{p=j+1}^{\infty}\gamma\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right\Vert _{r}&\leq &  \frac{1}{\mathfrak{m}}+\frac{ D}{\mathfrak{m}^2(1-\gamma \mathfrak{M})}.
Hence under assumptions of Theorem~\ref{th:mr}, there exists  a constant $\widetilde{C}_K < \infty$  such that for all $n \in \nset$ and $x \in \rset^d$,

@@

\PVar\left[\pi_{K,n}^{N}(f)\right] \leq \widetilde{C}_K  n^{-1} \gamma^{K-1}  \eqsp.

@@

We have
\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U(X_{k-1})\right)=\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right)+
\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right),
where empty products are equal $1$ by definition.
Hence
\left|\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U(X_{k-1})\right)\right|&\leq &\left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right)\prod_{k=j+1}^{p}\left |I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right|
&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}.

@@

Finally, using the Gaussian Poincare inequality, we show  that under  assumptions \eqref{eq:smooth-mu} and  \eqref{eq:main-ass-prod} it holds

@@

\right)\leq C_{K}\gamma^{-1}\Xi(x),\quad q=0,\ldots,n-1,

@@

\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2,

@@

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}. Due to Lemma~\ref{lem:var_bound},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
\]
It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
\[
\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
\]

@@

\right)\leq C\gamma^{-1} \Xi(x),\quad q=1,\ldots,n,

@@

\begin{lem}
\label{lem:var_poincare}
Under assumptions of Theorem~\ref{th:mr}, it holds
\begin{eqnarray}
\label{eq:lem-var}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  C_K\gamma^{-1}  \Xi(x), \quad q=1,\ldots,n,
\end{eqnarray}
where $C_K$ is a constant not depending on $x,$ $n$ and $\gamma.$
\end{lem}
\begin{proof}
The expression
$\sum_{p=1}^q \partial_{y_1}^K f(X_p^x)$
can be viewed as a deterministic function of
$x,Z_1,Z_{2},\ldots,Z_q$
$$
\sum_{p=1}^q  \partial_{y_1}^K f(X_p^x)
=F(x,Z_1,Z_{2},\ldots,Z_q).
$$
By the (conditional) Gaussian Poincar\'e inequality,
we have
$$
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
\le\mathsf E_x\left[
\|\nabla_Z F(x,Z_1,Z_{2},\ldots,Z_q)\|^2
\right],
$$
where $\nabla_Z F=(\partial_{Z_1} F,\ldots,\partial_{Z_q} F)$,
and $\|\cdot\|$ denotes the Euclidean norm.
Notice that \(\partial_{Z_j} F=\sqrt{\gamma}\,\partial_{y_j} F\) and
hence
$$
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=1}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
$$
It is straightforward to check that
$\partial_{y_j}\partial_{y_1}^K f(X_p^x)=0$
whenever $p<j$. Therefore, we get
\begin{equation}\label{eq:10062018a2}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=j}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
\end{equation}
Now fix $p$ and $j$, $p\ge j$, in $\{1,\ldots,q\}$.
By Fa\`a di Bruno's formula
\[
\partial_{y_{1}}^{K}f\left(X_{p}^x\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p}^x)\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}},
\]
where the sum is taken over all $K$-tuples of nonnegative integers
$(m_1,\ldots,m_K)$ satisfying
$$
1\cdot m_{1}+2\cdot m_{2}+\ldots+K\cdot m_{K}=K.
$$
Then
\begin{eqnarray*}
\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)
&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X^x_{p})\left[\partial_{y_{j}}X^x_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}^x}{k!}\right)^{m_{k}} \\
&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X^x_{p})
\sum_{s=1}^{K}\frac{m_{s}}{s!}
\left(\frac{\partial_{y_{1}}^{s}X^x_{p}}{s!}\right)^{m_{s}-1}
\\
&& \times\left[\partial_{y_{j}}\partial_{y_{1}}^{s}X^x_{p}\right]
\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}}.
\end{eqnarray*}
Using the bounds of
Lemmas \ref{lem:06062018a2} and~\ref{lem:06062018a3},
we obtain
\begin{eqnarray}
\label{eq:part-part}
\left|\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right|
\leq A_{K}\prod_{k=2}^{p}\alpha_{k}
\end{eqnarray}
with a suitable constant $A_{K}$.
Substituting this in~\eqref{eq:10062018a2},
we proceed as follows
\begin{align*}
\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
&\le
\gamma^2 A_{K}^{2}\sum_{j=1}^{q}
\mathsf{E}\left(\sum_{p=j}^{q}\prod_{k=2}^{p}\alpha_{k}\right)^{2}
\\
&\le
\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^2}
\mathsf{E}\sum_{j=1}^{q}
\left(\sum_{p=j+1}^{q+1}  \prod_{k=2}^{p} \alpha_{k}
\right)^{2}
\\
%&=
%\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
%\sum_{j=l}^{q} \gamma_{j}
%\prod_{k=l+1}^{j} \alpha_{k}^2
%\left(
%\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
%\right)^{2}
%\\
&\le
\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^3}
\mathsf{E}\sum_{j=1}^{q}
\prod_{k=1}^{j} \alpha_{k}
\left(\sum_{p=j+1}^{q+1}  \prod_{k=j+1}^{p} \alpha_{k}
\right)^{2}
\end{align*}
